### 2020-08-10 manuscript.tex
results update

\usepackage{mhchem}
\usepackage{mathtools}

@@

@@

% Results and Discussion can be combined.
\section*{Models} \label{sect:models}

@@

Let $\mathcal{T}$ be a binary rooted time tree with $N$ taxa. Let $L$ be the number of sites within the multiple sequence alignment $D$, and let $L_\text{eff}$ be the effective number of sites in the alignment (i.e. the number of unique patterns across all sites in the alignment). The posterior density of a phylogenetic model is described by

@@

\begin{align}
\label{eqn:adaptiveSampler}
	p(\omega_i) \propto \begin{cases} 1 & \text{ with probability } \Omega \\ \frac{1}{\mathbb{T}(\omega_i)} \sum\limits_{p \in \text{parameters}} \frac{1}{\sigma^2_p}  \sum\limits_{x \in \text{accepts}_i}  || x_p - x_p^\prime ||^2 & \text{ with probability } 1-\Omega \end{cases}
\end{align}

where $\Omega = 0.01$ allows any sub-operator to be sampled regardless of its performance.
The remaining terms are trained during the second and third phases: the cumulative computational runtime spent on each operator $\mathbb{T}(\omega_i)$, the sample variance $\sigma^2_p$ of each parameter $p$, and the sum-of-squares $\sum\limits_{x \in \text{accepts}_i} ||x_p - x_p^\prime||^2$, where $x_p$ and $x_p^\prime$ are the values of $p$ before and after each acceptance of a proposal made by $\omega_i$.

@@

Under \textbf{Equation \ref{eqn:adaptiveSampler}}, operators which effect larger changes on the parameters of interest, in shorter runtime, are sampled with greater probabilities. 
Division of the sum-of-squares term by the parameter variance $\sigma^2_p$ enables comparison between different parameters.

@@

\begin{table}[h!]
\centering
\begin{tabular}{|l p{4cm}|} 
 \hline
 Meta-operator & Operators \\
\hline
 \multirow{4}{*}{\texttt{AdaptiveOperatorSampler($\sigma$)}} & \texttt{FastClockScale($\sigma, \vec{\mathcal{R}}^{\,}$)}$^{*}$ \\ 
 & \texttt{RandomWalk($\sigma$)} \\
 & \texttt{Scale($\sigma$)} \\
 & \texttt{SampleFromPrior($\sigma$)} \\
 \hline
  \multirow{6}{*}{\texttt{AdaptiveOperatorSampler($\vec{\mathcal{R}}^{\,}$)}} & \texttt{ConstantDistance($\vec{\mathcal{R}}^{\,}, \mathcal{T}$)} \\ 
& \texttt{RandomWalk($\vec{\mathcal{R}}^{\,}$)} \\
& \texttt{Scale($\vec{\mathcal{R}}^{\,}$)}$^{*}$ \\
& \texttt{Interval($\vec{\mathcal{R}}^{\,}$)}$^{**}$  \\
& \texttt{Swap($\vec{\mathcal{R}}^{\,}$)} \\
& \texttt{SampleFromPrior($\vec{\mathcal{R}}^{\,}$)} \\
 \hline
   \multirow{2}{*}{\texttt{AdaptiveOperatorSampler(root)}} & \texttt{SimpleDistance($\vec{\mathcal{R}}^{\,}, \mathcal{T}$)} \\ 
&  \texttt{SmallPulley($\vec{\mathcal{R}}^{\,}, t$)}\\
 \hline
\end{tabular}
\caption{Summary of the three instances of \texttt{AdaptiveOperatorSampler} used under the \textit{real} and \textit{quant} parameterisations.
\texttt{AdaptiveOperatorSampler(root)} applies the root-targetting constant distance operators only \cite{zhang2020improving} while \texttt{AdaptiveOperatorSampler($\vec{\mathcal{R}}^{\,}$)} targets all rates and all nodes. 
The two meta-operators are weighted proportionally to the contribution of the root node to the total node count. 
  $^{*}$\textit{real} only. $^{**}$\textit{quant} only. }
\label{table:adaptiveSampling}
\end{table}

@@

The 48 operators generated by this pipeline are evaluated and compared in \textbf{\nameref{sect:results}}. Each operator is considered with and without a random walk on $t_D$ and thus there are 96 total settings.

@@

@@

Methodologies are benchmarked using one simulated and nine empirical datasets.
The latter were compiled \cite{lanfear2019Github} and partitioned \cite{lanfear2016partitionfinder} by Lanfear et al. as `benchmark alignments' (\textbf{Table \ref{table:datasets}}).
Each methodology is benchmarked across 5 replicates on each dataset, using an the Intel Xeon Gold 6138 CPU (2.00 GHz).
All methodologies use identical models and operator configurations, except where a difference is specified.

@@

 1  &  38  &  8  &  9.1  &  6.45  &  Seed plants (Ran 2018 \cite{Ran_2018}) \\

@@

7  &  78  &  8  &  3.4  &  3.1  &  Animals (Cannon 2016 \cite{Cannon_2016}) \\

8  &  80  &  1 &  10.0  &  0.9  &  \textit{Simulated data}  \\

9  &  94  &  4  &  2.2  &  1  &  Bees (Rightmyer 2013 \cite{Rightmyer_2013}) \\

10  &  106  &  1  &  0.8  &  0.5  &  Songbirds (Moyle 2016 \cite{Moyle_2016}) \\

 \hline
\end{tabular}
\caption{Datasets used during benchmarking, sorted in increasing order of taxa count $N$. Number of partitions $P$, total alignment length $L$, and number of patterns $L_\text{eff}$ are also specified.}
\label{table:datasets}
\end{table}

@@

We compared the cons, nocons, and adapt operator configurations for the \textit{real} and \textit{quant} parameterisations (\textbf{Table \ref{table:round1Results}}).
The nocons configuration contained the standard BEAST2 operator mix, while cons contained the operator mix used by Zhang and Drummond 2020 \cite{zhang2020improving}. 
The adapt mix combines the above operators, as well as the rudimentary \texttt{SampleFromPrior} operator, and learns the weights of each operator using  \texttt{AdaptiveOperatorSampler}.

\begin{table}[h!]
\centering
\begin{tabular}{|l l l l|} 
 \hline
Configuration & Operator & Weight & Parameterisations  \\
  \hline
 \multirow{6}{*}{nocons} & \texttt{RandomWalk($\mathcal{R}$)} & $10$ & \textit{real}  \\
 	& \texttt{Scale($\vec{\mathcal{R}}^{\,}$)} & $10$ & \textit{real} \\ 
 	& \texttt{Uniform($\vec{\mathcal{R}}^{\,}$)} & $10$ & \textit{quant} \\ 
 	& \texttt{Interval($\vec{\mathcal{R}}^{\,}$)} & $10$ & \textit{quant} \\ 
 	& \texttt{Swap($\vec{\mathcal{R}}^{\,}$)} & $10$ & \textit{real}, \textit{quant} \\ 
 	& \texttt{Scale($\sigma$)} & $10$ & \textit{real}, \textit{quant} \\ 
  \hline
   \multirow{10}{*}{cons} 
   & \texttt{ConstantDistance($\vec{\mathcal{R}}^{\,}, \mathcal{T}$)} & $20 \times \frac{2N-2}{2N-1}$ & \textit{real}, \textit{quant}  \\
   & \texttt{SimpleDistance($\vec{\mathcal{R}}^{\,}, \mathcal{T}$)} & $10 \times \frac{2N-2}{2N-1}$ & \textit{real}, \textit{quant}  \\
   & \texttt{SmallPulley($\vec{\mathcal{R}}^{\,}$)} & $10 \times \frac{2N-2}{2N-1}$ & \textit{real}, \textit{quant}  \\
   & \texttt{RandomWalk($\vec{\mathcal{R}}^{\,}$)} & $5$ & \textit{real}  \\
 	& \texttt{Scale($\vec{\mathcal{R}}^{\,}$)} & $2.5$ & \textit{real} \\ 
 	& \texttt{Uniform($\vec{\mathcal{R}}^{\,}$)} & $5$ & \textit{quant} \\ 
 	& \texttt{Interval($\vec{\mathcal{R}}^{\,}$)} & $2.5$ & \textit{quant} \\ 
 	& \texttt{Swap($\vec{\mathcal{R}}^{\,}$)} & $2.5$ & \textit{real}, \textit{quant} \\ 
 	& \texttt{FastClockScale($\sigma, \vec{\mathcal{R}}^{\,}$)} & $10$ & \textit{real}  \\ 
 	& \texttt{Scale($\sigma$)} & $10$ &  \textit{quant} \\ 
  \hline
     \multirow{3}{*}{adapt} & \texttt{AdaptiveOperatorSampler($\sigma$)} & $10$ & \textit{real}, \textit{quant} \\
 	& \texttt{AdaptiveOperatorSampler($\vec{\mathcal{R}}^{\,}$)} & $30 \times \frac{2N-2}{2N-1}$ & \textit{real}, \textit{quant} \\
 	& \texttt{AdaptiveOperatorSampler(root)} & $30 \times \frac{1}{2N-1}$ & \textit{real}, \textit{quant} \\
  \hline

\end{tabular}
\caption{Operator configurations for Round 1. In each configuration, the weight behind $\mathcal{R}$ sums to 30 and for $\sigma$ is equal to 10 within any one rate parameterisation. 
Operators which apply to either internal nodes or the root (but not both) are weighted according to leaf count $N$. The adapt operators are further broken down in \textbf{Table \ref{table:adaptiveSampling}}.} 
%The total weight of all operators across all parameters was 116 in each MCMC setup.
T%herefore a weight of 10 corresponds to sampling probability of 0.086, for example.
\label{table:round1Results}
\end{table}

These results show that, while nocons gives better performance than cons on smaller datasets (corresponding to low signal), and cons performs better on larger datasets (high signal), the adapt configuration gives the best performance overall.
This trend is observed for both the \textit{real} and the \textit{quant} parameterisations.
Therefore, the \texttt{AdaptiveOperatorSampler} operator will be included in all subsequent rounds in the tournament.

@@

We compared the three rate parameterisations described in \textbf{\nameref{sect:rateparams}}. 
The \textit{cat} operator configuration is described in \textbf{Table \ref{table:round2Results}}.
\textit{real} and \textit{quant} make use of constant distance tree operators \cite{zhang2020improving} and both use the \texttt{AdaptiveOperatorSampler} operator to learn clock model operator weights.
The three settings are validated in \textbf{\nameref{sect:WCSS_appendix}}.

\begin{table}[h!]
\centering
\begin{tabular}{|l l l|} 
 \hline
Configuration & Operator & Weight  \\
  \hline
 \multirow{4}{*}{\textit{cat}} 
 	& \texttt{RandomWalk($\vec{\mathcal{R}}^{\,}$)} & $10$ \\
 	& \texttt{Uniform($\vec{\mathcal{R}}^{\,}$)} & $10$  \\ 
 	& \texttt{Swap($\vec{\mathcal{R}}^{\,}$)} & $10$ \\ 
 	& \texttt{Scale($\sigma$)} & $10$ \\ 
  \hline
\end{tabular}
\caption{Operator configurations for \textit{cat} in Round 2. The configurations of \textit{real} (adapt) and \textit{quant} (adapt) are shown in \textbf{Table \ref{table:round1Results}}.}
\label{table:round2Results}
\end{table}

%\textbf{Fig. \ref{fig:parameterisationResults}} shows that the \textit{real 2006} performs considerably worse than any of the other settings. This is due to the poor sampling of the prior under this setting (i.e. low ESS of $p(\theta)$). The failure of \textit{real 2006} thus highlights the appeal of 1) the \textit{cat} or \textit{quant} parameterisations, both of which have trivial contributions to the prior density (i.e. uniform priors), and 2) the smarter operators used by \textit{real} (Zhang and Drummond 2020). Due to its computational burden, \textit{real 2006} was not benchmarked for all of the datasets in \textbf{Table \ref{table:datasets}}.

%Our results show that the \textit{quant} parameterisation yields the best performance with respect to effective samples per hour. \textit{quant} outperforms \textit{quant 2006}, suggesting that the constant distance operators are effective. Furthermore, \textit{quant} outperforms \textit{real} especially at sampling from the posterior and prior distributions (i.e. high ESS/hr for $p(\theta|D)$ and $p(\theta)$). This is most likely because of the uniform prior distribution of rate quantiles.

%Overall, \textit{quant} yields a median ESS/hr approximately 150 \% faster with respect to sampling the posterior probability, and approximately 30 \% faster with respect to sampling branch rates $r$ and clock standard deviation $\sigma$, compared with \textit{real}.

\begin{figure}[!h]
\includegraphics[width=\textwidth]{benchmarking/benchmarkingVM/ESS_round1.pdf}
\caption{\textbf{Rate parameterisation performance evaluation.} Comparison of ESS/hr (averaged across five independent MC3 analyses) with respect to relevant terms -- $P$: posterior density; $L$: likelihood, $p$: prior density, $r$: clock rate ESS averaged across all leaves, $\hat{r}$: branch rate mean, $v$: branch rate variance, $\sigma$: clock standard deviation, $\kappa$: HKY model transition-transversion ratio, $\lambda$: Yule model birth rate. $h$: tree height. Datasets are displayed in \textbf{Table \ref{table:datasets}}.  }
\label{fig:parameterisationResults}
\end{figure}

@@

\subsubsection*{Initial screening by acceptance rate on simulated data}

We selected the best operator variant by performing MCMC on 300 simulated datasets, where each MCMC employed all 96 NER/NERw variants. Simulated datasets have $N=30$ taxa and an alignment with $L \sim \text{Uniform}(10^2, 10^4)$ sites. The acceptance rate of each operator is compared to that of the null operator $\text{NER}\{\}$ (i.e. \texttt{Narrow Exchange}).

Although there are several operators which are tying for first place and behave equivalently, we selected the $\text{NER}\{\mathcal{D}_{AE}, \mathcal{D}_{BE}, \mathcal{D}_{CE}\}$ operator to proceed to the next round of optimisation.

@@

>>>>

\usepackage{commath}

\usepackage{mhchem}
\usepackage{tikz}
\usepackage{mathtools}

@@

\def\checkmark{\tikz\fill[scale=0.4](0,.35) -- (.25,0) -- (1,.7) -- (.25,.15) -- cycle;}

@@

% Results and Discussion can be combined.
\section*{Models and Methods} \label{sect:models}

@@

Let $\mathcal{T}$ be a binary rooted time tree with $N$ taxa. Let $L$ be the number of sites within the multiple sequence alignment $D$. 
The posterior density of a phylogenetic model is described by

@@

\begin{align}
\label{eqn:adaptiveSampler}
	p(\omega_i) \propto \begin{cases} 1 & \text{ with probability } \Omega \\ \frac{1}{\mathbb{T}(\omega_i)} \sum\limits_{p \in \text{POI}}  \sum\limits_{x \in \text{accepts}(\omega_i)}  \mathbb{D}(x_p, x_p^\prime) & \text{ with probability } 1-\Omega \end{cases}
\end{align}

where $\Omega = 0.01$ allows any sub-operator to be sampled regardless of its performance.
The parameters of interest (POI) may be either a set of numerical parameters (such as branch rates or node heights), or it may be the tree itself, but it cannot be both.
The distance between states $x_p$ and its (accepted) proposal $x_p^\prime$ with respect to parameter $p$ is determined by

\begin{align}
\label{eqn:distanceFunctions} 
	\mathbb{D}(x_p, x_p^\prime) = \begin{cases} \texttt{RF}(x_p, x_p^\prime)^2 & \text{ if } p \text{ is a tree} \\ 
	\frac{1}{|p|} \Big[ \frac{\norm{x_p - x_p^\prime}}{\sigma_p} \Big]^2    & \text{ if } p \text{ is numerical}  \end{cases}
\end{align}

where \texttt{RF} is the Robinson-Foulds tree distance \cite{robinson1981comparison}, and $|p|$ is the number of dimensions of numerical parameter $p$ (1 for $\sigma$, $2N-2$ for $\vec{\mathcal{R}}^{\,}$, and $2N-1$ for node heights $t$).
The remaining terms are trained during the second and third phases: the cumulative computational runtime spent on each operator $\mathbb{T}(\omega_i)$, the sample standard deviation $\sigma_p$ of each numerical parameter $p$, and the sum distances $\sum_x \mathbb{D}(x_p, x_p^\prime)$.

@@

Under \textbf{Equations \ref{eqn:adaptiveSampler} and \ref{eqn:distanceFunctions}}, operators which effect larger changes on the parameters of interest, in shorter runtime, are sampled with greater probabilities. 
Division of the squared distance by a parameters variance $\sigma^2_p$ enables comparison between different numerical parameters.

@@

\begin{table}[h!]
\centering
\begin{tabular}{|l l l|} 
 \hline
 Meta-operator & POI & Operators \\
\hline
 \multirow{4}{*}{\texttt{AdaptiveOperatorSampler($\sigma$)}} & \multirow{4}{*}{$\sigma$} & \texttt{FastClockScale($\sigma, \vec{\mathcal{R}}^{\,}$)} \\ 
 && \texttt{RandomWalk($\sigma$)}  \\
 && \texttt{Scale($\sigma$)}  \\
 && \texttt{SampleFromPrior($\sigma$)}  \\
 \hline
  \multirow{6}{*}{\texttt{AdaptiveOperatorSampler($\vec{\mathcal{R}}^{\,}$)}} & \multirow{6}{*}{$\vec{\mathcal{R}}^{\,}, t$} & \texttt{ConstantDistance($\vec{\mathcal{R}}^{\,}, \mathcal{T}$)}   \\ 
&& \texttt{RandomWalk($\vec{\mathcal{R}}^{\,}$)}  \\
&& \texttt{Scale($\vec{\mathcal{R}}^{\,}$)}   \\
&& \texttt{Interval($\vec{\mathcal{R}}^{\,}$)}   \\
&& \texttt{Swap($\vec{\mathcal{R}}^{\,}$)}  \\
&& \texttt{SampleFromPrior($\vec{\mathcal{R}}^{\,}$)} \\
 \hline
   \multirow{2}{*}{\texttt{AdaptiveOperatorSampler(root)}} & \multirow{2}{*}{$\vec{\mathcal{R}}^{\,}, t$} & \texttt{SimpleDistance($\vec{\mathcal{R}}^{\,}, \mathcal{T}$)}  \\ 
&&  \texttt{SmallPulley($\vec{\mathcal{R}}^{\,}, t$)}  \\
 \hline
\end{tabular}
\caption{Summary of the three instances of \texttt{AdaptiveOperatorSampler} and their parameters of interest (POI).
Note that different operators are applicable to different substitution rate parameterisations (\textbf{Table \ref{table:kernels}}). 
\texttt{AdaptiveOperatorSampler(root)} applies the root-targeting constant distance operators only \cite{zhang2020improving} while \texttt{AdaptiveOperatorSampler($\vec{\mathcal{R}}^{\,}$)} targets all rates and all nodes heights $t$. 
The two meta-operators are weighted proportionally to the contribution of the root node to the total node count. }
\label{table:adaptiveSampling}
\end{table}

@@

\subsubsection*{Screening operators by acceptance rate on simulated data}

We selected the best NER operator variant by performing MCMC on 300 simulated datasets, where each MCMC employed all 96 NER/NERw variants. 
Simulated datasets have $N=30$ taxa and an alignment with $L \sim \text{Uniform}(10^2, 10^4)$ sites. 
The acceptance rate of each operator (which in this case is equivalent to the average \texttt{RF} distance) is compared to that of the null operator $\text{NER}\{\}$ (i.e. \texttt{NarrowExchange}).

Although there are several operators which behaved equivalently during this initial screening process, we selected $\text{NER}\{\mathcal{D}_{AE}, \mathcal{D}_{BE}, \mathcal{D}_{CE}\}$ to proceed to benchmarking (\textbf{\nameref{sect:results}}).
Due to the apparent sensitivity of NER operators to the dataset, we introduce the adaptive operator \texttt{AdaptiveOperatorSampler(NER)} which allows the operator scheme to fall back on the standard \texttt{NarrowExchange} in the event of $\text{NER}\{\mathcal{D}_{AE}, \mathcal{D}_{BE}, \mathcal{D}_{CE}\}$ performing poorly (\textbf{Table \ref{table:adaptiveNER}}).

\begin{table}[h!]
\centering
\begin{tabular}{|l l l|} 
 \hline
 Meta-operator & POI & Operators \\
\hline
 \multirow{2}{*}{\texttt{AdaptiveOperatorSampler(NER)}} & \multirow{2}{*}{$\mathcal{T}$} & NER$\{  \}$ \\ 
 && NER$\{ \mathcal{D}_{AE}, \mathcal{D}_{BE}, \mathcal{D}_{CE} \}$ \\
 \hline
\end{tabular}
\caption{The adaptive NER operator. 
The Robinson-Foulds distance between trees before and after every proposal accept is used to train the operator weights. In the special case of NER proposals, the \texttt{RF} distance is always equal to 1. }
\label{table:adaptiveNER}
\end{table}

@@

\subsection*{Operator configurations} \label{sect:operatorSchemes}

Throughout the article, we have introduced four new operators. 
These are summarised in \textbf{Table \ref{table:newOperators}}.

In \textbf{Table \ref{table:operatorSchemes}}, we define all operator configurations which are benchmarked throughout \textbf{\nameref{sect:results}}.

\begin{table}[h!]
\centering
\begin{tabular}{|l l l l l l|} 
 \hline
Configuration & Operator & Weight & \textit{real} & \textit{cat} & \textit{quant}  \\
  \hline
 \multirow{6}{*}{nocons} & \texttt{RandomWalk($\mathcal{R}$)} & $10$ & \checkmark &&  \\
 	& \texttt{Scale($\vec{\mathcal{R}}^{\,}$)} & $10$ & \checkmark && \\ 
 	& \texttt{Uniform($\vec{\mathcal{R}}^{\,}$)} & $10$ & & \checkmark & \checkmark \\ 
 	& \texttt{Interval($\vec{\mathcal{R}}^{\,}$)} & $10$ &  & & \checkmark \\ 
 	& \texttt{Swap($\vec{\mathcal{R}}^{\,}$)} & $10$ & \checkmark & \checkmark & \checkmark \\ 
 	& \texttt{Scale($\sigma$)} & $10$ & \checkmark & \checkmark & \checkmark \\ 
  \hline
   \multirow{10}{*}{cons} 
   & \texttt{ConstantDistance($\vec{\mathcal{R}}^{\,}, \mathcal{T}$)} & $20 \times \frac{2N-2}{2N-1}$ & \checkmark & & \checkmark  \\
   & \texttt{SimpleDistance($\vec{\mathcal{R}}^{\,}, \mathcal{T}$)} & $10 \times \frac{2N-2}{2N-1}$ & \checkmark & & \checkmark  \\
   & \texttt{SmallPulley($\vec{\mathcal{R}}^{\,}$)} & $10 \times \frac{2N-2}{2N-1}$ & \checkmark & & \checkmark  \\
   & \texttt{RandomWalk($\vec{\mathcal{R}}^{\,}$)} & $5$ & \checkmark &&  \\
 	& \texttt{Scale($\vec{\mathcal{R}}^{\,}$)} & $2.5$ & \checkmark && \\ 
 	& \texttt{Uniform($\vec{\mathcal{R}}^{\,}$)} & $5$ & &&\checkmark \\ 
 	& \texttt{Interval($\vec{\mathcal{R}}^{\,}$)} & $2.5$ & &&\checkmark\\ 
 	& \texttt{Swap($\vec{\mathcal{R}}^{\,}$)} & $2.5$ & \checkmark & & \checkmark  \\ 
 	& \texttt{FastClockScale($\sigma, \vec{\mathcal{R}}^{\,}$)} & $10$ & \checkmark  && \\ 
 	& \texttt{Scale($\sigma$)} & $10$ &  &&\checkmark \\ 
  \hline
     \multirow{3}{*}{adapt} & \texttt{AdaptiveOperatorSampler($\sigma$)} & $10$ & \checkmark & \checkmark & \checkmark  \\
 	& \texttt{AdaptiveOperatorSampler($\vec{\mathcal{R}}^{\,}$)} & $30 \times \frac{2N-2}{2N-1}$ & \checkmark & & \checkmark  \\
 	& \texttt{AdaptiveOperatorSampler($\vec{\mathcal{R}}^{\,}$)} & $30$ &  & \checkmark & \\
 	& \texttt{AdaptiveOperatorSampler(root)} & $30 \times \frac{1}{2N-1}$ & \checkmark & & \checkmark  \\

  \hline

\end{tabular}
\caption{Operator configurations and the substitution rate parameterisations which each operator are applicable to. Within each configuration (and substitution rate parameterisation), the weight behind $\mathcal{R}$ sums to 30 and the weight of $\sigma$ is equal to 10. 
There is no cons (\textit{cat}) configuration because the parameterisation is not compatible the constant distance operators.
Operators which apply to either internal nodes or the root (but not both) are weighted according to leaf count $N$. The adapt operators are further broken down in \textbf{Table \ref{table:adaptiveSampling}}.} 
%The total weight of all operators across all parameters was 116 in each MCMC setup.
T%herefore a weight of 10 corresponds to sampling probability of 0.086, for example.
\label{table:operatorSchemes}
\end{table}

@@

Methodologies are benchmarked using one simulated and eight empirical datasets.
The latter were compiled \cite{lanfear2019Github} and partitioned \cite{lanfear2016partitionfinder} by Lanfear et al. as `benchmark alignments' (\textbf{Table \ref{table:datasets}}).
Each methodology is benchmarked across 5 replicates on each dataset, using an the Intel Xeon Gold 6138 CPU (2.00 GHz).
All methodologies use identical models and operator configurations, except where a difference is specified.

@@

 1  &  38  &  16  &  15.5  &  10.5  &  Seed plants (Ran 2018 \cite{Ran_2018}) \\

@@

7  &  80  &  1 &  10.0  &  0.9  &  \textit{Simulated data}  \\

8  &  94  &  4  &  2.2  &  1  &  Bees (Rightmyer 2013 \cite{Rightmyer_2013}) \\

9  &  106  &  1  &  0.8  &  0.5  &  Songbirds (Moyle 2016 \cite{Moyle_2016}) \\

 \hline
\end{tabular}
\caption{Datasets used during benchmarking, sorted in increasing order of taxa count $N$. Number of partitions $P$, total alignment length $L$, and number of site patterns $L_\text{eff}$ in the alignment are also specified.}
\label{table:datasets}
\end{table}

@@

We compared the cons, nocons, and adapt operator configurations for the \textit{real} and \textit{quant} parameterisations (\textbf{Table \ref{table:operatorSchemes}}). 
The nocons configuration contained the standard BEAST2 operator mix, while cons contained the operator mix used by Zhang and Drummond 2020 \cite{zhang2020improving}. 
The adapt mix combines the above operators, as well as the simple-but-bold \texttt{SampleFromPrior} operator, and learns the weights of each operator using  \texttt{AdaptiveOperatorSampler}.
We also tested the adapt (\textit{cat}) configuration to determine whether \textit{cat} can be further improved by learning the underlying weights behind \texttt{RandomWalk}, \texttt{Swap}, and \texttt{SampleFromPrior}.

These results show that nocons performs better than cons on small datasets while cons performs better on large datasets (\textbf{Fig. \ref{fig:round1Results}} and \textbf{\nameref{sect:SR_appendix}}). 
This is an expected result and is true for both the \textit{real} and the \textit{quant} parameterisations.
The adapt configuration dramatically improved mixing for \textit{real}, by finding the right balance between these two extremes.
However the \texttt{AdaptiveOperatorSampler} operator did not benefit for either \textit{cat} or \textit{quant}, suggesting that the default operator weighting scheme was sufficient in either case.
Overall, the \texttt{AdaptiveOperatorSampler} operator will be included in all subsequent rounds in the tournament.

\begin{figure}[!h]
\includegraphics[width=\textwidth]{benchmarking/benchmarkingVM/ESS_round1_real.pdf}
\caption{\textbf{Round 1: benchmarking the \texttt{AdaptiveOperatorSampler} operator.} Top left, top right, bottom left: each plot compares the ESS/hr ($\pm 2$ standard errors) across two operator configurations with respect to relevant terms -- $L$: likelihood, $p$: prior density, $r$: clock rate ESS averaged across all leaves, $\sigma$: clock standard deviation, $h$: tree height, and $\kappa$: HKY model transition-transversion ratio.
Bottom right: the effect of sequence length $L$ on operator weights learned by \texttt{AdaptiveOperatorSampler}. Both sets of results are fit by a logistic regression curve.
 The benchmark datasets are displayed in \textbf{Table \ref{table:datasets}}. The \emph{cat} and \emph{quant} evaluations are presented in \nameref{sect:SR_appendix}.  }
\label{fig:round1Results}
\end{figure}

@@

We compared the three rate parameterisations described in \textbf{\nameref{sect:rateparams}}. 
adapt (\textit{real}) and adapt (\textit{quant}) both employ constant distance tree operators \cite{zhang2020improving} and both use the \texttt{AdaptiveOperatorSampler} operator to learn clock model operator weights.
Clock model operators weights are also learned in the adapt (\textit{cat}) configuration.

This experiment revealed that the \textit{real} parameterisation greatly outperforms both \textit{cat} and \textit{quant} on most datasets (\textbf{Fig. \ref{fig:round2Results}}).
This disparity is strongest for large datasets.
In the most extreme case, $r$ and $\sigma$ both mixed around $70 \times$ faster on the 15.5 kb Seed plant dataset by \cite{Ran_2018} for \textit{real} than they did for \textit{cat}.
The advantages in using constant distance operators would likely be even stronger for larger $L$.
Overall, we have determined that \textit{real}, and its associated operators, makes the best parameterisation covered here and will proceed to the next rounds of the benchmarking.

\begin{figure}[!h]
\includegraphics[width=\textwidth]{benchmarking/benchmarkingVM/ESS_round2.pdf}
\caption{\textbf{Round 2: benchmarking substitution rate parameterisations.} Top left, top right, bottom left: the adapt (\textit{real}), adapt (\textit{cat}), and adapt (\textit{quant}) configurations were compared. See \textbf{Fig. \ref{fig:round1Results}} caption for figure notation. Bottom right: comparison of the mean tip rate ESS/hr as a function of sequence length. }
\label{fig:round2Results}
\end{figure}

@@

@@

\paragraph*{S3 Appendix.}
\label{sect:SR_appendix}
{\textbf{Supplementary results.} Presentation and interpretation of further benchmarking results.
